Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-03-01 paper.tex
Deadlock -> Starvation

\begin{abstract}
  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding deadlock. Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\section{Introduction}
Scan is one of the most ubiquitous operations in GPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$. \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{speed-of-light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite starvation and will never complete. In practice, it is easy to demonstrate this highly undesirable behavior.

@@

We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving SOL performance without requiring forward-progress guarantees (FPG). Our implementation targets portability across modern GPU hardware and software platforms. It is implemented in the WebGPU Shading Language (WGSL)---an apex-level shading language that transpiles to backend API languages---ensuring compatibility with all backend graphics APIs: Vulkan, Direct3D, and Metal. Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.

@@

In any \emph{Chained Scan}, once a workgroup computes its local scan, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles. This process, which we refer to as \emph{Chained Propagation}, defines the set of inter-workgroup techniques that facilitate dependency resolution. In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume. This approach depends on a forward-progress guarantee of the GPU's execution model. Given such a guarantee, the predecessor will eventually produce a result and broadcast it. On devices lacking such a guarantee, the workgroup may wait indefinitely, resulting in deadlock.

@@

\subsubsection{Chained Scan, Decoupled Lookback}
Introduced by Merrill and Garland, the \emph{Decoupled Lookback} technique refines \emph{Chained Propagation} by separating reduction communication from dependency resolution, mitigating bottlenecks caused by propagation latency. In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction. Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback. However, granting each workgroup control over its dependency resolution does not eliminate reliance on \emph{progress guarantees}, as workgroups must still spin on tiles that have yet to post their reductions, risking deadlock.

@@

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. Furthermore, its single-kernel design and serial ordering make it suitable for in-situ operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging. However, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple's GPUs do not. Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Progress Guarantees and Their Impact on \emph{Chained Scan}}
A \emph{forward-progress guarantee} ensures that all occupant subgroups---of possibly different workgroups---eventually make progress toward termination, preventing indefinite starvation. Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees. Instead, Sorensen et al.\ systematically proposed and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
\begin{itemize}
  \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
  \item \textbf{Heterogeneous System Architecture~\cite[Section~2.11]{HSASysArch12} (HSA)}: The workgroup with the lowest ID that has not yet terminated.
  \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
\end{itemize}
Their 2021 findings~\cite{sorensen2021} demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior hypothesis~\cite{sorensen2018}.

@@

\subsection{Earlier Attempts at Portability}
Our first attempt at implementing a \emph{Decoupled Lookback} scan without relying on FPG was \emph{Scalar Fallback}. Each time a workgroup spins waiting on the reduction from a preceding tile, it advances a \emph{fallback} operation by processing one scalar element from that tile. This eliminates the possibility of deadlock because the number of such operations is bounded; the wait will eventually terminate, possibly at the expense of doing redundant work.

@@

  \item \textbf{Lookback:}
        \begin{itemize}
          \item The first thread queries the state of \emph{current tile}.
                \begin{itemize}
                  \item If the tile is in the \emph{Ready} state, the workgroup ingests the value and sets \emph{current tile} to the preceding tile, repeating this step until reaching a tile that is \emph{Not Ready}.
                  \item If the tile is in the \emph{Inclusive} state, the workgroup ingests the value and proceeds directly to step~4.
                  \item If the tile is in the \emph{Not Ready} state, the first thread spins for a limited number of iterations, defined by the \emph{Maximum Spin Count}.
                        \begin{itemize}
                          \item If this count is exceeded, it exits the lookback loop and proceeds to step~3, leaving the control flag \emph{Locked} to signal to the workgroup that a fallback is required. Additionally, it copies \emph{current tile} into shared memory to broadcast the fallback tile identity to the workgroup.
                          \item Otherwise, it continues spinning until the tile transitions to a \emph{Ready} or \emph{Inclusive} state. Without FPG, \emph{Decoupled Lookback} risks spinning indefinitely and deadlocking at this stage.
                        \end{itemize}
                \end{itemize}
        \end{itemize}

>>>>

\begin{abstract}
  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding \emph{starvation}. Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\section{Introduction}
Scan is one of the most ubiquitous operations in GPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$. \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{speed-of-light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite \emph{starvation} and will never complete. In practice, it is easy to demonstrate this highly undesirable behavior.

@@

We present \emph{Decoupled Fallback}, a fully portable \emph{Chained Scan} capable of achieving SOL performance without requiring forward-progress guarantees (FPG). Our implementation targets portability across modern GPU hardware and software platforms. It is implemented in the WebGPU Shading Language (WGSL)---an apex-level shading language that transpiles to backend API languages---ensuring compatibility with all backend graphics APIs: Vulkan, Direct3D, and Metal. Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.

@@

In any \emph{Chained Scan}, once a workgroup computes its local scan, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles. This process, which we refer to as \emph{Chained Propagation}, defines the set of inter-workgroup techniques that facilitate dependency resolution. In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume. This approach depends on a forward-progress guarantee of the GPU's execution model. Given such a guarantee, the predecessor will eventually produce a result and broadcast it. On devices lacking such a guarantee, the workgroup may continuously spin without making progress, resulting in \emph{starvation}---consuming execution resources that would otherwise allow the blocking tile to complete its reduction and stalling the kernel indefinitely.

@@

\subsubsection{Chained Scan, Decoupled Lookback}
Introduced by Merrill and Garland, the \emph{Decoupled Lookback} technique refines \emph{Chained Propagation} by separating reduction communication from dependency resolution, mitigating bottlenecks caused by propagation latency. In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction. Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback. However, granting each workgroup control over its dependency resolution does not eliminate reliance on \emph{progress guarantees}, as workgroups must still spin on tiles that have yet to post their reductions, risking starvation.

@@

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. Furthermore, its single-kernel design and serial ordering make it suitable for in-situ operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging. However, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple's GPUs do not. Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite starvation, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Progress Guarantees and Their Impact on \emph{Chained Scan}}
A \emph{forward-progress guarantee} ensures that all occupant subgroups---of possibly different workgroups---are eventually given execution and make progress toward termination, preventing indefinite starvation. Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees. Instead, Sorensen et al.\ systematically proposed and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
\begin{itemize}
  \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
  \item \textbf{Heterogeneous System Architecture~\cite[Section~2.11]{HSASysArch12} (HSA)}: The workgroup with the lowest ID that has not yet terminated.
  \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
\end{itemize}
Their 2021 findings~\cite{sorensen2021} demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior hypothesis~\cite{sorensen2018}.

@@

\subsection{Earlier Attempts at Portability}
Our first attempt at implementing a \emph{Decoupled Lookback} scan without relying on FPG was \emph{Scalar Fallback}. Each time a workgroup spins waiting on the reduction from a preceding tile, it advances a \emph{fallback} operation by processing one scalar element from that tile. This eliminates the possibility of starvation because the number of such operations is bounded; the wait will eventually terminate, possibly at the expense of doing redundant work.

@@

  \item \textbf{Lookback:}
        \begin{itemize}
          \item The first thread queries the state of \emph{current tile}.
                \begin{itemize}
                  \item If the tile is in the \emph{Ready} state, the workgroup ingests the value and sets \emph{current tile} to the preceding tile, repeating this step until reaching a tile that is \emph{Not Ready}.
                  \item If the tile is in the \emph{Inclusive} state, the workgroup ingests the value and proceeds directly to step~4.
                  \item If the tile is in the \emph{Not Ready} state, the first thread spins for a limited number of iterations, defined by the \emph{Maximum Spin Count}.
                        \begin{itemize}
                          \item If this count is exceeded, it exits the lookback loop and proceeds to step~3, leaving the control flag \emph{Locked} to signal to the workgroup that a fallback is required. Additionally, it copies \emph{current tile} into shared memory to broadcast the fallback tile identity to the workgroup.
                          \item Otherwise, it continues spinning until the tile transitions to a \emph{Ready} or \emph{Inclusive} state. Without FPG, \emph{Decoupled Lookback} risks spinning indefinitely at this stage.
                        \end{itemize}
                \end{itemize}
        \end{itemize}
